Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-23 paper.tex
Add in bit about annotations
Closes #71
Closes #72
Closes #73

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]

@@

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices [citations?].
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks requiring negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiences of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data among interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of various fields are determined by the \emph{maximum}
dimension among all variants. Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider problems of ongoing updates
(i.e, as new samples and their variants are added).
Thirdly, Zarr is not suitable for handling unstructured
string data, and for example likely not work well as a 
basis for storing annotation data [CITES].

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or require storage larger 
than uncompressed VCF~\citep{al2023critical}.

@@

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices [citations?].
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks require negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiences of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data in a large ecosystem of interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends 
quadratically on the number of alleles (although practical
solutions have been suggested that can be 
adapted~\cite{poterba2024scalable}).
Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider how samples (and the 
corresponding novel variants) should be added.
Thirdly, Zarr is not suitable for handling 
unstructured data often included in VCF INFO 
fields, and would likely be a poor choice
for representing variant 
annotations, for example.
